---
title: "10X Multiome - 5 timepoints - CMO channel 1 QC"
output: html_document
---

# 10X multiome 5 timepoints channel 1

```{r}
suppressPackageStartupMessages(library(Seurat))
suppressPackageStartupMessages(library(here))
suppressPackageStartupMessages(library(ggplot2))

ec_diff_palette = c(
  "#C6C7C7",
  "#A8B1D6",
  "#EBBC9E",
  "#FBC1C3",
  "#F7999C"
)

cmo_counts_path <- here(
  "data",
  "cmo_counts",
  "channel1",
  "counts_unfiltered"
)
```

# Load assignments from Dulguun

```{r}
cmo_assignments_old <- read.delim(
  here(
    "data",
    "barcode_mapping_multiome_dulguun.dedup.tsv"
  ),
  header = FALSE,
  stringsAsFactors = FALSE
)
colnames(cmo_assignments_old) = c("barcode", "old_assignment") 

# Keep only barcodes for channel 1 (-1_1)
cmo_assignments_old_ch1 <-  cmo_assignments_old %>%
  filter(grepl("-1_1$", barcode)) %>%
  mutate(
    barcode_new = sub("-1_1$", "", barcode)
  )

cmo_assignments_old_ch1$old_assignment_clean <- recode(
  cmo_assignments_old_ch1$old_assignment,
  "d0" = "d0_ipsc",
  "d1" = "d1_ps",
  "d2" = "d2_meso",
  "d3" = "d3_ec",
  "d4" = "d4_ec",
  .default = cmo_assignments_old_ch1$old_assignment
)

head(cmo_assignments_old_ch1)
```

# Barplot old assignments

```{r}
plot_df_diff_old <- cmo_assignments_old_ch1 %>%
  count(old_assignment_clean)

ggplot(plot_df_diff_old, aes(x = old_assignment_clean, y = n, fill=old_assignment_clean)) +
  geom_bar(stat = "identity") +
  theme_minimal() +
  labs(
    title = "Classification of CMO Data - From Dulguun",
    subtitle = "Filtered Singlets only",
    x = "Differentiation stage",
    y = "Number of Barcodes"
    ) +
  scale_fill_manual(values = ec_diff_palette)
```

# Load CMO counts into Seurat

```{r}
mat <- Matrix::readMM(file.path(cmo_counts_path, "cells_x_features.mtx"))
barcodes <- readLines(file.path(cmo_counts_path, "cells_x_features.barcodes.txt"))
genes <- readLines(file.path(cmo_counts_path, "cells_x_features.genes.txt"))
gene_names <- readLines(file.path(cmo_counts_path, "cells_x_features.genes.names.txt"))

# Seurat expects genes as rows and barcodes as columns, so we need to transpose the matrix
counts <- t(mat)
rownames(counts) <- genes
colnames(counts) <- barcodes
# Create Seurat object
seu.cmo.raw <- CreateSeuratObject(
  counts = counts,
  assay = "CMO"
)
DefaultAssay(seu.cmo.raw) <- "CMO"

# Create the metadata
  cmo_ids <- paste0("CMO", sprintf("%02d", 1:25))
  cmo_meta <- data.frame(
    cmo_id = cmo_ids,
    condition = rep(c("d0_ipsc", "d1_ps", "d2_meso", "d3_ec", "d4_ec"),each = 5),
    replicate = rep(1:5, times = 5),
    row.names = cmo_ids
  )
# Add the metadata to the object
seu.cmo.raw[["CMO"]] <- AddMetaData(
  object = seu.cmo.raw[["CMO"]],
  metadata = cmo_meta[rownames(seu.cmo.raw[["CMO"]]), , drop = FALSE]
)

head(seu.cmo.raw)
```

```{r}
# Merge the old assignments with the Seurat metadata
meta <- seu.cmo.raw@meta.data

meta$old_assignment <- cmo_assignments_old_ch1$old_assignment_clean[
  match(rownames(meta), cmo_assignments_old_ch1$barcode_new)
]

meta$old_assignment[is.na(meta$old_assignment)] <- "Unassigned"

seu.cmo.raw@meta.data <- meta

seu.cmo.raw@meta.data$old_assignment %>%
  table() %>%
  prop.table()
```

# CMO classes to barcodes using Seurat::HTODemux

Seurat::HTODemux will classify cells as "Singlet", "Doublet", or "Negative" based on the CMO counts. The positive.quantile parameter can be adjusted to set the threshold for calling a cell positive for a given CMO.

```{r}
library(here)
flt_bc = read.delim(here("results/channel1/filtered_barcodes.txt"), header=FALSE)
flt_bc
```


```{r warning=FALSE}
# keep cells/barcodes with old assignment
#cells_keep <- rownames(seu.cmo.raw@meta.data)[
#  seu.cmo.raw@meta.data$old_assignment != "Unassigned"
#]
colnames(seu.cmo.raw) <- paste0(
  colnames(seu.cmo.raw),
  "_10x_5timepoints_channel1"
)

cells_keep <- intersect(flt_bc$V1, colnames(seu.cmo.raw))

seu.cmo <- subset(
  seu.cmo.raw,
  cells = cells_keep
)

DefaultAssay(seu.cmo) <- "CMO"

cmo_counts <- GetAssayData(seu.cmo, assay = "CMO", layer = "counts")

cells_nonzero <- colnames(cmo_counts)[Matrix::colSums(cmo_counts) > 0]

seu.cmo <- subset(
  seu.cmo,
  cells = cells_nonzero
)

seu.cmo <- NormalizeData(
  seu.cmo,
  assay = "CMO",
  normalization.method = "CLR",
  margin = 2
)

seu.cmo <- HTODemux(
  seu.cmo,
  assay = "CMO",
  positive.quantile = 0.99
)
```
```{r}
hist(seu.cmo)
```

```{r}
# Compute the fraction of Singlet, Doublet, and Negative classifications
seu.cmo$CMO_classification.global %>%
  table() %>%
  prop.table()
```

```{r}
seu.cmo@meta.data$hash.ID <- factor(
  seu.cmo@meta.data$hash.ID,
  levels = c(cmo_ids, "Doublet", "Negative")
)

seu.cmo$hash.ID %>%
  table()
```


```{r}
# Barplot of the classifications
plot_df <-data.frame(
  classification = names(table(seu.cmo$hash.ID)),
  count = as.numeric(table(seu.cmo$hash.ID))
)
# Remove negative classifications
plot_df <- plot_df %>%
  filter(classification != "Negative" & classification != "Doublet")

plot_df <- merge(
  plot_df,
  data.frame(classification = cmo_ids, differentiation = rep(c("d0_ipsc", "d1_ps", "d2_meso", "d3_ec", "d4_ec"), each = 5)),
  by.x = "classification",
  by.y = "classification",
  all.x = TRUE
)

plot_df$classification <- factor(plot_df$classification, levels = c(cmo_ids, "Doublet"))

ggplot(plot_df, aes(x = classification, y = count, fill=differentiation)) +
  geom_bar(stat = "identity") +
  theme_minimal() +
  labs(
    title = "HTODemux Classification of CMO Data",
    subtitle = "Singlets only",
    x = "CMO Hash ID",
    y = "Number of Barcodes"
    ) +
  scale_x_discrete(guide = guide_axis(angle = 45)) +
  scale_fill_manual(values = ec_diff_palette)
```


```{r}
# Total by differentiation stages
plot_df_diff <- plot_df %>%
  group_by(differentiation) %>%
  summarise(count = sum(count))

ggplot(plot_df_diff, aes(x = differentiation, y = count, fill=differentiation)) +
  geom_bar(stat = "identity") +
  theme_minimal() +
  labs(
    title = "HTODemux Classification of CMO Data",
    subtitle = "Singlets only",
    x = "Differentiation stage",
    y = "Number of Barcodes"
    ) +
  scale_x_discrete(guide = guide_axis(angle = 45)) +
  scale_fill_manual(values = ec_diff_palette)
```
# Compare with old assignments

```{r}
compare_df <- seu.cmo@meta.data %>%
  select(new_assignment = hash.ID, old_assignment = old_assignment)

# Map CMO to differentiation stages
compare_df <- compare_df %>%
  left_join(
    data.frame(
      hash.ID = cmo_ids,
      differentiation = rep(c("d0", "d1", "d2", "d3", "d4"), each = 5)
    ),
    by = c("new_assignment" = "hash.ID")
  ) %>%
  mutate(
    match = old_assignment == differentiation,
  )

temp = as.data.frame(table("new" = compare_df$differentiation, "old" = compare_df$old_assignment, useNA = "ifany"))
ggplot(temp, aes(x = old, y = new)) +
  geom_tile(fill = NA) +
  geom_text(aes(label = Freq)) +
  theme_minimal() +
  labs(x = "Old", y = "New")
```

```{r}
write.table(seu.cmo@meta.data, file = here("data", "cmo_counts", "channel1", "cmo_assignments_metadata.tsv"), sep = "\t", quote = FALSE, row.names = TRUE)

```

```{r}
FeatureScatter(seu.cmo, feature1 = "CMO18", feature2 = "CMO24")
```

```{r}
Seurat::RidgePlot(seu.cmo, cols=c(rep(ec_diff_palette, each=5),"black"),group.by = "hash.ID",
,features = c("CMO01"))
```


```{r}
positive_quantiles <- c(0.90, 0.95, 0.975, 0.99, 0.995, 0.999)

results <- lapply(positive_quantiles, function(q) {
  obj <- subset(seu.cmo.raw, cells = cells_keep)
  DefaultAssay(obj) <- "CMO"
  
  obj <- NormalizeData(
    obj,
    assay = "CMO",
    normalization.method = "CLR",
    margin = 2,
    verbose = FALSE
  )
  
  obj <- HTODemux(
    obj,
    assay = "CMO",
    positive.quantile = q,
    verbose = FALSE
  )
  
  pred_condition <- cmo_meta[obj$hash.ID, "condition"]
  
  old <- obj$old_assignment
  
  keep <- !is.na(pred_condition) & old != "Unassigned"
  
  data.frame(
    positive.quantile = q,
    n_cells = length(old),
    n_compared = sum(keep),
    agreement = mean(pred_condition[keep] == old[keep]),
    singlets = sum(obj$CMO_classification.global == "Singlet"),
    doublets = sum(obj$CMO_classification.global == "Doublet"),
    negatives = sum(obj$CMO_classification.global == "Negative")
  )
})

results_df <- bind_rows(results)
results_df
```

```{r}
results_df %>%
  arrange(desc(agreement), desc(singlets))
```

```{r}
best_q <- results_df$positive.quantile[which.max(results_df$agreement)]
best_q
```

```{r}
obj <- subset(seu.cmo.raw, cells = cells_keep)
DefaultAssay(obj) <- "CMO"

obj <- NormalizeData(obj, assay = "CMO", normalization.method = "CLR", margin = 2)
obj <- HTODemux(obj, assay = "CMO", positive.quantile = best_q)

obj$pred_condition <- cmo_meta[obj$hash.ID, "condition"]

table(
  old = obj$old_assignment,
  new = obj$pred_condition,
  useNA = "ifany"
)
```

```{r}
cmo_mat <- GetAssayData(seu.cmo, assay = "CMO", layer = "counts")

top_call <- apply(cmo_mat, 2, function(x) {
  names(which.max(x))
})

seu.cmo$CMO_argmax <- top_call
```

```{r}
seu.cmo$CMO_argmax %>%
  table()
```

```{r}
seu.cmo$CMO_simple <- apply(cmo_mat, 2, function(x) {
  names(which.max(x))
})
```

```{r}
seu.cmo$timepoint <- sub("_.*", "", cmo_meta[seu.cmo$hash.ID, "condition"])
seu.cmo$timepoint_argmax <- sub("_.*", "", cmo_meta[seu.cmo$CMO_simple, "condition"])
table(
  old = seu.cmo$old_assignment,
  new = seu.cmo$timepoint_argmax
)
```

```{r}
seu.cmo@meta.data[seu.cmo$old_assignment=="d2" & seu.cmo$timepoint_argmax=="d3",]
```

```{r}
barplot(seu.cmo[["CMO"]]$counts[,"AAGAATCAGCTATATG"])
```

